Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-03-11 young_pp_discs.tex
added 15 deg casa sims to body, re-working main text

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \smallskip

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of \citet{Boley2007} and  \citet{Boley2009}. The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$au in $x$, $y$ and $z$. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history. In the dense conditions of the protoplanetary disc could be encouraging the dust grains to agglomerate and increase the size of the ice mantle through freeze out of volitiles. This could lead to changes in the dust opacity, however we have not attempted to include these effects in our model.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$au$^2$.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \cite{Caselli2012}). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the LAMDA (Leiden Atomic and Molecular DAtabase) database (Sch\"oier et al. 2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime2.eps}%Lime_grid3.eps
 \caption{A plot of the points selected by the gridding process and the paths down which photons can propagate colour coded by the density distribution (in m$^{-3}$, as used in LIME). The points are more concentrated at small radii and in the most dense regions.}
 \label{grid}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope, the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 \citep{Caselli2012}, scaled so the maximum is 1$\times$10$^{-8}$ (assuming, for simplicity, that photodissociation and freeze-out affect these two species in a similar way).For H$_2$CO, a step profile was considered, with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{-9}$ inside it \citep{Young2004}. For OCS we adopted a constant abundance of 1.9$\times$10$^{-9}$ \citep{Ren2011}. The C$^{17}$O abundance profile in the envelope follows the CO profile calculated for L1544 by KC2010, scaled by the 17O/16O elemental abundance. \smallskip

@@

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and the disc is inclined at 30$^\circ$ to edge on. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity and intensity weighted velocity maps (figures \ref{sim_all} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope. This can be seen in some PV diagrams  (Figure \ref{pvs}) as the strong absorption feature at all positions around zero velocity. Moment 1 maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

@@

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (figure \ref{sim_all}). As OCS is not seen in outflows (e.g. \citealt{Stanke2007}, \citealt{VDTak2003}) it can be used to view rotation of the central part of the disc without risk of contamination from emission from shocked material along the outflow. The OCS line, which has the highest upper energy level among the selected transitions, traces the hottest and densest regions of the disc. As a result, the innermost spiral structure is best seen in OCS (26 - 25), as shown in figure \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, because of the presence of the protostar and other hot material within the central hole, not considered in our simple model, we expect absorption to also be observed toward the central region of the disc.\smallskip

@@

Figure \ref{continuum} Shows the 1mm continuum emission and beam size in this configuration of ALMA.

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$ line overlaid on the 1$\,$mm continuum emission. From this it can be seen that the absorption features trace the disc spiral structure whilst the line emission emanates from gaps in the outer disc, where there is less continuum to be absorbed.\smallskip

@@

Figure \ref{mom0_maps} shows the simulated integrated intensity maps, which demonstrates how lines from different species allow observations to trace different regions of the disc, from the OCS line tracing the innermost regions, to the HCO$^+$ line showing the outer regions which are less bright in the continuum. ALMA has enough resolution and sensitivity to see spiral structure clearly in continuum emission but also in CO isotopologues and marginally in H$_2$CO.\smallskip

@@

\section*{Acknowledgements}
PC and TWH acknowledge support of successive rolling grants awarded by the UK Science and Technology Funding Council. 
TAD and JDI acknowledge studentships from the Science and Technology Facilities Council of the United Kingdom (STFC).

\bibliographystyle{apalike} 
\bibliography{new_bib}

@@

\section{Grid Construction} \label{sec:gridding}

In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their density and molecular density compared against a reference point in order to decide if the point is to be used in the grid or not. Candidate grid points are selected at random in cylindrical coordinates, linearly spaced in z and $\phi$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, otherwise another r, $\phi$, z co-ordinate is selected and this becomes the candidate point. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all the scales while ensuring that the majority of points are located in the inner disc where the density is higher. 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See figure \ref{points} for an example of the points distribution in r, z. \smallskip

>>>>

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). Although L1544 does not appear to contain a central protostar and disc itself, other similar dense cores such as L1521F contain a central object which appears to be a low mass protostar and disc \citep{Bourke2006} whilest having around half the mass of L1544. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \smallskip

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of \citet{Boley2007} and  \citet{Boley2009}. The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms, rotating around a 1$\,$M$_\odot$ mass protostar (which is not included in the model). H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$au in $x$, $y$ and $z$. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history. In the dense conditions of the protoplanetary disc could be encouraging the dust grains to agglomerate and increase the size of the ice mantle through freeze out of volitiles. This could lead to changes in the dust opacity, however we have not attempted to include these effects in our model.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$au$^2$.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \cite{Caselli2012}). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the Leiden Atomic and Molecular DAtabase (LAMDA) (Sch\"oier et al. 2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime2.eps}%Lime_grid3.eps
 \caption{A plot of the points selected by the gridding process and the paths down which photons can propagate for points in the central r,$\theta$ plane. The points are colour coded by the density distribution (in m$^{-3}$, as used in LIME) and are more concentrated at small radii and in the most dense regions. The circular high density region in the centre is the disc model and is 128$\,$au in diameter.}
 \label{grid}
\end{figure}

\subsection{Grid construction} \label{subsec:gridding}
In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their density and molecular density compared against a reference point in order to decide if the point is to be used in the grid or not. Candidate grid points are selected at random in cylindrical coordinates, linearly spaced in z and $\phi$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, otherwise another r, $\phi$, z co-ordinate is selected and this becomes the candidate point. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all the scales while ensuring that the majority of points are located in the inner disc where the density is higher. 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See figure \ref{points} for an example of the points distribution in r, z. \smallskip

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope, the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 \citep{Caselli2012}, scaled so the maximum is 1$\times$10$^{-8}$ (assuming, for simplicity, that photodissociation and freeze-out affect these two species in a similar way).For H$_2$CO, a step profile was considered, with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{-9}$ inside it \citep{Young2004}. For OCS we adopted a constant abundance of 1.9$\times$10$^{-9}$ \citep{Ren2011}. The C$^{17}$O abundance profile in the envelope follows the CO profile calculated for L1544 by KC2010, scaled by the 17O/16O elemental abundance. These estimates for the abundances in the envelope are simplistic, however, as seen in section \ref{alma_predictions}, the envelope contribution to the line is very narrow, and can be spectrally disentangled from the disc contribution.\smallskip

@@

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc). The inlination angle was varied from 15$^\circ$ to 75$^\circ$ to edge on, we will consider the 30$^\circ$ case as a 'typical' inclination with the 15$^\circ$ inclination discussed in section \ref{alma_predictions}, the other inclinations simulated are shown in appendix \ref{other_inc}. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity and intensity weighted velocity maps (figures \ref{sim_all} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope. This can be seen in some PV diagrams  (Figure \ref{pvs}) as the strong absorption feature at all positions around zero velocity. Moment 1 maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

@@

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (figure \ref{sim_all}). As OCS is not seen in outflows (e.g. \citealt{Stanke2007}, \citealt{VDTak2003}) it can be used to view rotation of the central part of the disc without risk of contamination from emission from shocked material along the outflow. The OCS line, which has the highest upper energy level among the selected transitions, traces the hottest and densest regions of the disc. As a result, the innermost spiral structure is best seen in OCS (26 - 25), as shown in figure \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, because of the presence of the protostar and other hot material within 2$\,$au of the protostar, not considered in our simple model, we expect absorption to also be observed toward the central region of the disc.\smallskip

@@

Figure \ref{continuum} Shows the 1mm continuum emission and beam size in this configuration of ALMA. The spiral structure of the disc is clearly visible at this inclination and will be clearer the closer the disc is to being face on. \smallskip

Figure \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$ line overlaid on the 1$\,$mm continuum emission. From this it can be seen that the absorption features trace the disc spiral structure whilst the line emission emanates from gaps in the outer disc, where there is less continuum to be absorbed. The central pannel is domminated by the envelope contribution, causing stron absorption through out the disc. The other pannels show the line following a rotation profile distorted by the presence of the spiral arms. \smallskip

@@

Figure \ref{mom0_maps} shows the simulated integrated intensity maps, which demonstrates how lines from different species allow observations to trace different regions of the disc, from the OCS line tracing the innermost regions, to the HCO$^+$ line showing the outer regions which are less bright in the continuum. ALMA has enough resolution and sensitivity to see spiral structure clearly in continuum emission but also in CO isotopologues and marginally in H$_2$CO. The different molecules can be seen to be tracing different regions of the disc\smallskip

\begin{figure}
 \includegraphics[width=63mm]{Figures/sim/casa_C17O_15deg_all.eps}

 \caption{CASA simulation of the C17O 3$\rightarrow$2 line inclided at 15$^\circ$ to edge on {\bf Top:} integrated intensity map. {\bf bottom:} position velocity diagram. In the position velocity diagram the spiral structures can be seen as narrow finger-like structures extending from the origin.}
 \label{15deg}
\end{figure}

%\begin{figure}
%  \centering
%  \includegraphics[width=42mm]{Figures/sim/casa_C17O_15deg_PV.eps}

%  \caption{CASA simulation of the position velocity diagram of the C17O 3$\rightarrow$2 line inclided at 15$^\circ$ to edge on.}
%  \label{15deg_pv}
%\end{figure}

@@

\section*{Acknowledgements}
PC and TWH acknowledge support of successive rolling grants awarded by the UK Science and Technology Funding Council. 
TAD and JDI acknowledge studentships from the Science and Technology Facilities Council of the United Kingdom (STFC).
The research leading to these results has received funding from the European Union FP7-2011 under grant agreement no 284405.

\bibliographystyle{mn2e} 
\bibliography{new_bib}

@@
